Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-10-05 Manuscript/manuscript.tex
Included Pavel's changes until the end of section III. C.

%In order to validate the structural properties of the newly developed ECC-lipids model
%we first compared our simulation results without any ions to
%NMR order parameters measurements and x-ray scattering form factors 
%(Fig.~\ref{simVSexpNOions}). 
The x-ray scattering form factor (Fig.~\ref{simVSexpNOions}) and
the area per lipid (Table~\ref{tab:apls}) of POPC bilayer simulated
with the ECC-lipid model are in good agreement with the experimental results,
as well as with the Lipid14 model. Thus, we conclude that the lipid bilayer dimensions
are well captured by the developed ECC-lipid model. 
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.

@@

\subsection{Response of POPC head group to bound charge}

@@

Before proceeding to the ion binding affinity studies, we quantify
the response of headgroup order parameters to the amount of 
bound charge by using mixtures of monovalent cationic surfactants
and POPC~\cite{scherer89}. The amount of bound charge per PC 
in these systems is given by the molar fraction of cationic 
surfactants, because essentially all surfactants locate in the 
lipid bilayers. Experimental data for these systems can be used to validate 
the sensitivity of lipid headgroup order parameters
to the amount of bound charge in simulations.

@@

The headgroup order parameter changes with increasing amount of 
cationic surfactant dihexadecyldimethylammonium bromide 
is compared between experiments~\cite{scherer89} and simulations in
Fig. \ref{OrderParameterCHANGESsurf}. 
The observed order parameter decrease is simulations and experiments
can be approximated to be linear at least with the mole fractions
below $\sim$30\%, as expected from Eq. \ref{OPchangeEQ}.
The slope is, however, too steep in Lipid14 model indicating that 
the head group order parameters are too sensitive to a bound charge.
The ECC-lipids model gives a slope in very good agreement with experiments
for the $\alpha$ segment, while the slope is slightly
underestimated for the $\beta$ segment.
%This has to be taken into account when interpreting the
%head group order parameter response to CaCl$_2$ concentration in Lipid14 model.
%observed in  Ref. \citenum{catte16}, see also Fig. \ref{fig:delta_ordPar_NaCl}.
%The present comparison of head group order parameter changes 
%suggests that the observed overestimated response of all models to 
%increasing CaCl$_2$ concentration in \cite{catte16} 
%can be explained at least in part by a high sensitivity of the head group order parameter response
%to the bound charge.
\todo{SAMULI: We could calculate the slopes from simulations, but I am not sure if
we would actually learn anything useful from this.}

The headgroup P-N vector angle with respect to the membrane normal
is also shown as a function of cationic surfactant mole fraction
in Fig.~\ref{OrderParameterCHANGESsurf}.
The headgroup orients more towards the water phase with increasing
amount of bound cations, as previously reported in Ref.~\citenum{seelig87}. 
The effect is more pronounced in Lipid14 than 
in ECC-lipids model, which is in line with the order parameter 
results and the reduced charge--dipole interactions in the ECC-lipid model.
The response of $\alpha$-order parameter to bound positive charge
in ECC-lipid model is in good agreement with experiments.
The model can be thus used to study changes of lipid P-N vector
in varying conditions.

@@

\subsection{Cation binding affinity in POPC}

@@

The binding affinity of aqueous cations in lipid 
bilayers can be measured by using the headgroup 
order parameters, because they decrease proportionally  
to the bound positive charge~\cite{seelig87,catte16}. 
The headgroup order parameter responses 
to aqueous NaCl and CaCl$_2$ concentrations 
are shown in Fig.~\ref{fig:delta_ordPar_CaCl} and S\ref{fig:delta_ordPar_NaCl}
from experiments (DPPC \cite{akutsu81} 
and POPC \cite{altenbach84}) and different simulation models for POPC.

@@

Negligible changes of the headgroup order parameters are measured
with submolar concentrations of NaCl due to the very low affinity of Na$^+$
in PC bilayers \cite{akutsu81}. While Na$^+$ binding and the related 
headgroup order parameter changes were overestimated in
almost all the available simulation models, the low affinity
and negligible order parameter changes were reproduced by Lipid14 
model when simulated with \AA{}qvist ions~\cite{catte16}.
However, the same combination of force field parameters
overestimated the headgroup order parameter response to CaCl$_2$ 
concentration, which was the case also in all other models tested in
Ref.~\citenum{catte16}.
Using ion model by Dang et al.~\cite{smith94,chang1999,dang2006} or
ECC-ions~\cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}
with more realistic bulk behaviour did not improve the results
for the CaCl$_2$ interactions with Lipid14 model, as seen 
in Figs. \ref{fig:delta_ordPar_CaCl} and S\ref{fig:ordPars_waterModels} (in SI), respectively.
\todo{Add OP-response of Lipid14+ECC-ions plot in SI}.
The results support the conclusion of the previous work \cite{catte16}
that improvements also in lipid models are needed to 
correctly describe divalent cation binding in PC bilayers.

Significant improvement can be achieved by using the ECC
approach also for lipids. The headgroup order parameter changes
as a function of CaCl$_2$ concentration from ECC-lipid model with ECC-ions
show a good agreement with experiments in Fig.~\ref{fig:delta_ordPar_CaCl} and S\ref{fig:delta_ordPar_NaCl}.
As discussed in previous section, the model gives also
a good agreement with experiments for
the headgroup response to bound charge.
Thus, the model can used for more detailed analysis of the
binding affinity.

The binding affinities are quantified by using
the water density profiles along membrane normal shown
in Fig.~\ref{fig:cacl-dens}. The density profiles show
larger Ca$^{2+}$ density peak in lipid headgroup region for
Lipid14 model with Dang and ECC-ions than for the ECC-lipid model.
The relative surface excess calculated from Eq.~\ref{surfexcess} gives 
$\Gamma_i^w = 0.07 \pm 0.01 nm^{-2}$ for the ECC-lipid model,
which is significantly smaller than $\Gamma_i^w = 0.13 \pm 0.01 nm^{-2}$ for
Lipid14 with \AA{}qvist and $\Gamma_i^w = 0.3 \pm 0.03 nm^{-2}$ with Dang ions.

\todo{Below analysis is done in a stupid way to get some idea. I would fine it useful to
  do this analysis by using the density profiles, but it is not necessary.
  The reason why I want to do something like this, is that it would be useful to have
  some easier way to estimate the correct binding affinity than the electrometer concept,
  which is quite tedious to apply in practise (simulations with different concenrations,
  cationic surfactant check, etc.). I wuold like to be able to estimate much faster from a
  simulation if the binding affinity is reasonable or not. This may not be the correct way for
that anyway.}
Rough estimates for the free energy difference between bound and unbound cation
are given by
\begin{equation}
  \Delta G={\rm k_b T}\log(\frac{p_o}{p_i}),
\end{equation}
where $p_o$ and $p_i$ are estimated from the Ca$^{2+}$densities in bulk water and
in the maximum density in bilayer, respectively. The density profiles in
Fig. \ref{fig:cacl-dens} give $\sim 0.8$ k$_{\rm b}$T for the free energy difference
between bound and unbound Ca$^{2+}$ ions in ECC-lipid model and $\sim 1.4$ k$_{\rm b}$T
in Lipid14 with \AA{}qvist.

@@

Since the lipid headgroup order parameter responses to the amount of bound charge
and to the aqueous ion concentrations are both in good agreement with experiments
in the ECC-lipid model with ECC-ion parameters, we consider the Na$^+$ and Ca$^{2+}$
binding affinities to be realistic in this model. On the other hand, the Ca$^{2+}$
binding affinity is overestimated by Lipid14 model when simulated with all the
tested ion models. Similar conclusions were previously made based only on the headrogup
order parameter data with aqueous cations~\cite{catte16}. However,
the discrepancies with experiments in previous work could partly arise also from
the inaccurate sensitivity of the headgroup to bound charge. Here we quantify this
effect and conclude that the improvement due to ECC is partly, but not completely,
caused by more realistic headgroup sensitivity to bound charge.
This indicates that the issue should be carefully considered also when
the electrometer concept is used to compare ion binding between experiments
and simulations wiht other models.

>>>>

%In order to validate the structural properties of the newly developed ECC-lipids model
%we first compared our simulation results without any ions to
%NMR order parameters measurements and x-ray scattering form factors 
%(Fig.~\ref{simVSexpNOions}). 
The x-ray scattering form factor (Fig.~\ref{simVSexpNOions}) and
the area per lipid (Table~\ref{tab:apls}) of POPC bilayer simulated
with the ECC-lipid model are in good agreement with the experimental results,
as well as with the Lipid14 model. Thus, we conclude that the lipid bilayer dimensions
are well captured by the developed ECC-lipid model. 
Areas per lipid of ECC-lipids with various other water models are documented in SI, table~\ref{tab:apls_si}. 
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.

@@

\subsection{Response of POPC head groups to bound charge}

@@

Before proceeding to the ion binding affinity, we quantify
the response of headgroup order parameters to the amount of 
bound charge by using mixtures of monovalent cationic surfactants
and POPC~\cite{scherer89}. The amount of bound charge per PC 
in these systems is given by the molar fraction of cationic 
surfactants, because essentially all surfactants locate in the 
lipid bilayers. Experimental data for these systems can be used to validate 
the sensitivity of lipid headgroup order parameters
to the amount of bound charge in simulations.

@@

The headgroup order parameter changes with an increasing amount of 
the cationic surfactant dihexadecyldimethylammonium bromide 
is compared between experiments~\cite{scherer89} and simulations in
Fig. \ref{OrderParameterCHANGESsurf}. 
The observed order parameter decrease in simulations and experiments
can be approximated to be linear at least for mole fractions
below $\sim$30\%, as expected from Eq. \ref{OPchangeEQ}.
The slope is, however, too steep in the Lipid14 model indicating that 
the head group order parameters are too sensitive to a bound charge.
The ECC-lipids model gives a slope in very good agreement with experiments
for the $\alpha$ segment, while the slope is slightly
underestimated for the $\beta$ segment.
%This has to be taken into account when interpreting the
%head group order parameter response to CaCl$_2$ concentration in Lipid14 model.
%observed in  Ref. \citenum{catte16}, see also Fig. \ref{fig:delta_ordPar_NaCl}.
%The present comparison of head group order parameter changes 
%suggests that the observed overestimated response of all models to 
%increasing CaCl$_2$ concentration in \cite{catte16} 
%can be explained at least in part by a high sensitivity of the head group order parameter response
%to the bound charge.
\todo{SAMULI: We could calculate the slopes from simulations, but I am not sure if
we would actually learn anything useful from this.}

The headgroup P-N vector angle with respect to the membrane normal
is also shown as a function of cationic surfactant mole fraction
in Fig.~\ref{OrderParameterCHANGESsurf}.
The headgroup orients more towards the water phase with an increasing
amount of bound cations, as previously reported in Ref.~\citenum{seelig87}. 
The effect is more pronounced in Lipid14 than 
in ECC-lipids model, which is in line with the order parameter 
results and the reduced charge--dipole interactions in the latter model.
The response of $\alpha$-order parameter to bound positive charge
in ECC-lipid model is in good agreement with experiments.
The model can be thus used to study changes of lipid P-N vector
in varying conditions.

@@

\subsection{Cation binding affinities to POPC}

@@

The binding affinities of aqueous cations to lipid 
bilayers can be measured using the headgroup 
order parameters, because they decrease proportionally  
to the bound positive charge~\cite{seelig87,catte16}. 
The headgroup order parameter responses 
to increasing aqueous CaCl$_2$ concentrations 
from experiments (DPPC \cite{akutsu81} 
and POPC \cite{altenbach84}) and different simulation models for POPC 
are shown in Fig.~\ref{fig:delta_ordPar_CaCl}.
Responses to NaCl concentrations are in SI, Fig.~S\ref{fig:delta_ordPar_NaCl}.

@@

Negligible changes of the headgroup order parameters are measured
with submolar concentrations of NaCl due to the very low affinity of Na$^+$
in PC bilayers \cite{akutsu81}. While Na$^+$ binding and the related 
headgroup order parameter changes were overestimated in
almost all the available simulation models, the low affinity
and negligible order parameter changes were reproduced by Lipid14 
model when simulated with \AA{}qvist ions~\cite{catte16}.
However, the same combination of force field parameters
overestimated the headgroup order parameter response to CaCl$_2$ 
concentration, which was the case also in all other models tested in
Ref.~\citenum{catte16}.
Using the ion model by Dang et al.~\cite{smith94,chang1999,dang2006} or
ECC-ions~\cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}
with more realistic bulk behaviour did not improve the results
for the CaCl$_2$ interactions with a POPC bilayer described by the Lipid14 model, as seen 
in Figs. \ref{fig:delta_ordPar_CaCl} and S\ref{fig:ordPars_waterModels} (in SI), respectively.
\todo{Add OP-response of Lipid14+ECC-ions plot in SI}.
The results support the conclusion of the previous work \cite{catte16}
that improvements also in lipid models are needed to 
correctly describe divalent cation binding to PC bilayers.

Significant improvement can be achieved using the ECC
approach also for lipids. The headgroup order parameters change
as a function of CaCl$_2$ concentration for the ECC-lipid model with ECC-ions
exhibit a good agreement with experiments in Fig.~\ref{fig:delta_ordPar_CaCl} and S\ref{fig:delta_ordPar_NaCl}.
As discussed in previous section, the model gives also
a good agreement with experiments for
the headgroup response to bound charge.
%Thus, the model can used for more detailed analysis of the binding affinity.

The binding affinities are quantified using
the water density profiles along the membrane normal shown
in Fig.~\ref{fig:cacl-dens}. The density profiles show
a larger Ca$^{2+}$ density peak in lipid headgroup region for
the Lipid14 model with Dang and ECC-ions than for the ECC-lipid model.
The relative surface excess calculated from Eq.~\ref{surfexcess} is 
$\Gamma_i^w = 0.07 \pm 0.01 nm^{-2}$ for the ECC-lipid model,
which is significantly smaller than $\Gamma_i^w = 0.13 \pm 0.01 nm^{-2}$ for
Lipid14 with \AA{}qvist and $\Gamma_i^w = 0.3 \pm 0.03 nm^{-2}$ with Dang ions.

\todo{Below analysis is done in a stupid way to get some idea. I would fine it useful to
  do this analysis by using the density profiles, but it is not necessary.
  The reason why I want to do something like this, is that it would be useful to have
  some easier way to estimate the correct binding affinity than the electrometer concept,
  which is quite tedious to apply in practise (simulations with different concenrations,
  cationic surfactant check, etc.). I wuold like to be able to estimate much faster from a
  simulation if the binding affinity is reasonable or not. This may not be the correct way for
that anyway.}
Rough estimates for the free energy difference between bound and unbound cation
are given by
\begin{equation}
  \Delta G = -{\rm k_b T}\log(\frac{p_i}{p_o}),
\end{equation}
where $p_o$ and $p_i$ are estimated from the Ca$^{2+}$densities in bulk water and
in the maximum density in bilayer, respectively. The density profiles in
Fig. \ref{fig:cacl-dens} give $\sim 0.8$ k$_{\rm b}$T for the free energy difference
between bound and unbound Ca$^{2+}$ ions in ECC-lipid model and $\sim 1.4$ k$_{\rm b}$T
in Lipid14 with \AA{}qvist.

@@

Since the lipid headgroup order parameter responds to the amount of bound charge
and to the aqueous ion concentrations are both in good agreement with experiments
in the ECC-lipid model with ECC-ion parameters, we consider the Na$^+$ and Ca$^{2+}$
binding affinities to be realisticlly described in this model. In contrast, the Ca$^{2+}$
binding affinity is overestimated by Lipid14 model when simulated with all the
tested ion models. Similar conclusions were previously made based only on the headrogup
order parameter data with aqueous cations~\cite{catte16}. However,
the discrepancies with experiments in previous work could partly arise also from
the inaccurately described sensitivity of the headgroup to bound charge. Here we quantify this
effect and conclude that the improvement due to ECC is to a large extent, but not entirely,
caused by a more realistic headgroup sensitivity to bound charge.
This indicates that the issue should be carefully considered also when
the electrometer concept is used to compare ion binding between experiments
and simulations with other models.
